\Large\bf FlexibleSUSY --- A spectrum generator generator for supersymmetric models

### 2014-06-03 doc/paper.tex
Revert "use proper shell command"
This reverts commit b1babd2c67c2e10aabd79ed65529321b6be7cd9b.

Conflicts:

	doc/paper.tex
doc/paper.tex: minor corrections

  \begin{abstract}
   We introduce \fs, a \mathematica package which generates a fast,
   precise C++ spectrum generator for any SUSY model specified by the
   user.  The generated code is designed with both speed and
   modularity in mind, making it easy to adapt and extend with new
   features. The user specifies the model supplying the
   superpotential, gauge structure and particle content in a \sarah
   model file and provides specific the boundary conditions in a
   separate \fs model file. \fs makes use of the existing \sarah
   package to obtain self energies, tadpole corrections,
   renormalisation group equations(RGEs) and tree level mass and
   electroweak symmetry breaking conditions for the specified model. These are
   translated into C++ code and combined  with numerical
   routines for solving the RGEs, EWSB conditions and simultaneously
   solving for the spectrum conssistent with user specified boundary
   conditions.  The modular structure of the generated code allows for
   individual components to be replaced with an alternative if
   available. \fs has been carefully designed to grow as
   alternative solvers and calculators are added.
  \end{abstract}

@@

\section{Program Summary}
\noindent{\em Program title:} \fs\\ {\em Program obtainable from:}
         {\tt http://flexiblesusy.hepforge.org/}\\ {\em Distribution
           format:}\/ tar.gz\\ {\em Programming language:} {\tt
           C++}\\ {\em Computer:}\/ Personal computer\\ {\em Operating
           system:}\/ Tested on Linux 3.x\\ {\em Word size:}\/ 64
         bits\\ {\em External routines:}\/ SARAH 4.0.4, Boost library,
         Eigen, LAPACK\\ {\em
           Typical running time:}\/ 0.1-0.3 seconds per parameter
         point.\\ {\em Nature of problem:}\/ Determining the mass
         spectrum and mixings for any supersymmetric model. The
         generated code must find simultaneous solutions to
         constraints which are specified at two or more different
         renormalisation scales, which are connected by
         renormalisation group equations forming a large set of
         coupled differential equations. \\ {\em Solution method:}\/
         Nested iterative algorithm and numerical minimisation of the
         Higgs potential.  \\ {\em Restrictions:} The couplings must
         remain perturbative at all scales between the highest and
         lowest boundary condition.  \fs~ assumes that all couplings
         of the model are real (i.e.\ $CP-$conserving). Due to the
         modular nature of the generated code adaption and extension
         to overcome restrictions in scope is quite straight forward.

\newpage
\section{Introduction}
Supersymmetry(SUSY) provides the only non-trivial way to extend the
space-time symmetries of the Poincar\'e
group \cite{Coleman:1967ad,Haag:1974qh}, leading many to suspect that
SUSY may be realised in nature in some form. In particular
supersymmetric extensions of the standard model where SUSY is broken
at the TeV scale have been proposed to solve the hierarchy
problem \cite{Weinberg:1975gm, Weinberg:1979bn, Gildener:1976ai,
  Susskind:1978ms, 'tHooft:1980xb}, allow gauge coupling
unification \cite{Langacker:1990jh, Ellis:1990wk, Amaldi:1991cn,
  Langacker:1991an, Giunti:1991ta} and predict a dark matter candidate
which can fit the observed relic
density \cite{Goldberg:1983nd,Ellis:1983ew}.  Such models have also
been used for baryogenesis or leptogensis to solve the
matter-anti-matter asymmetry of the universe and have been considered
as the low energy effective models originating from string
theory.

@@

Recent experimental developments have also increased the relevancy of
such a tool. From the recent $7$ TeV and $8$ TeV runs at the Large
Hadron Collider (LHC) there have been two important developments.
First low energy signatures expected from such models, such as the
classic jets plus missing transverse energy signature, have not been
observed, substantially raising the lower limit on sparticle masses
(see e.g.~\cite{Aad:2013wta,Chatrchyan:2014lfa}). No other signature
of beyond the standard model(BSM) physics has been observed, leaving
the fundamental questions which motivated BSM physics
unanswered. Secondly ATLAS and CMS discovered\cite{ATLAS:2012ae,
  Chatrchyan:2012tx} a light Higgs of $125$ GeV, within the mass range
that could be accommodated in the MSSM but requiring stops which are
significantly heavier than both the direct collider limits and
indirect limits that appears in constrained models from the
significantly higher limits on first and second generation squarks.

These motivate both exploring non-minimal SUSY models which ameliorate
the naturalness problems by, for example, raising the tree level Higgs
mass, and models developed with a fresh perspective, based on other
considerations.  In both cases exploration of such models can be aided
if it is possible to quickly create a fast spectrum generator.
Currently there is only option for this, a SPheno-like FORTRAN code
which can be generated from
\sarah\cite{Staub:2010ty,Staub:2009bi,Staub:2010jh,Staub:2012pb,Staub:2013tta}.

@@

\paragraph{Alternative boundary condition problem solvers}

@@

\fs can be downloaded as a gzipped tar file from the
\url{http://flexiblesusy.hepforge.org/}.  To download and install
version 1.0.0 run:
%
\begin{lstlisting}[language=bash]
$ wget https://www.hepforge.org/archive/flexiblesusy/FlexibleSUSY-1.0.0.tar.gz
$ tar -xf FlexibleSUSY-1.0.0.tar.gz
$ cd FlexibleSUSY-1.0.0/
\end{lstlisting}%% $
%
A CMSSM spectrum generator can be created with the following three
commands:
%
\begin{lstlisting}[language=bash]
$ ./createmodel --name=MSSM
$ ./configure --with-models=MSSM
$ make
\end{lstlisting}%% $
%
The first command creates the model directory \code{models/MSSM/}
together with a CMSSM model file.  The \code{configure} script checks
the system requirements and creates the \code{Makefile}.  See
\code{./configure --help} for more options.  Executing \code{make}
will start \mathematica to generate the spectrum generator and compile
it.  The resulting executable can be run like this:
\begin{lstlisting}[language=bash]
$ ./models/MSSM/run_MSSM.x \
    --slha-input-file=model_files/MSSM/LesHouches.in.MSSM
\end{lstlisting}%% $
%
When executed, the spectrum generator tries to find a set of \DRbar\
model parameters consistent with all CMSSM boundary conditions for the
parameter point given in the SLHA input file
\code{model_files/MSSM/LesHouches.in.MSSM}.  Afterwards, the pole mass
spectrum and mixing matrices are calculated and written to the command
line in SLHA format \cite{Skands:2003cj,Allanach:2008qq}.  For the
parameter point given in the above example the calculated pole mass
spectrum reads
%
\begin{lstlisting}
Block MASS
   1000021     1.15236966E+03   # Glu
   1000024     3.85774334E+02   # Cha_1
   1000037     6.50460073E+02   # Cha_2
        25     1.14766149E+02   # hh_1
        35     7.06792640E+02   # hh_2
        37     7.11388516E+02   # Hpm_2
        36     7.06523105E+02   # Ah_2
   1000012     3.51856376E+02   # Sv_1
   1000014     3.53042556E+02   # Sv_2
   1000016     3.53046504E+02   # Sv_3
   1000022     2.03889780E+02   # Chi_1
   1000023     3.85760714E+02   # Chi_2
   1000025     6.36544884E+02   # Chi_3
   1000035     6.50133768E+02   # Chi_4
   1000001     9.66656018E+02   # Sd_1
   1000003     1.00983181E+03   # Sd_2
   1000005     1.01651873E+03   # Sd_3
   2000001     1.01653005E+03   # Sd_4
   2000003     1.06089534E+03   # Sd_5
   2000005     1.06090238E+03   # Sd_6
   1000011     2.22570305E+02   # Se_1
   1000013     2.29864536E+02   # Se_2
   1000015     2.29888846E+02   # Se_3
   2000011     3.61946671E+02   # Se_4
   2000013     3.61950866E+02   # Se_5
   2000015     3.63136031E+02   # Se_6
   1000002     8.09787818E+02   # Su_1
   1000004     1.01454197E+03   # Su_2
   1000006     1.01981109E+03   # Su_3
   2000002     1.02015269E+03   # Su_4
   2000004     1.05807759E+03   # Su_5
   2000006     1.05808168E+03   # Su_6
\end{lstlisting}

@@

By default, \fs creates three default boundary conditions for each
SUSY model:
%
\begin{itemize}
\item The \emph{high-scale constraint} is indended to set boundary
  conditions on the model parameters at some very high scale, e.g.\
  the GUT scale $M_X$.  The high-scale is defined by the value given
  in the variable \code{HighScale}.  In the model file in
  \secref{Sec:modfile} it is defined to be the unification scale $M_X$
  where $g_1(M_X) = g_2(M_X)$.  The model parameters which, are set in
  the $apply()$ function at the high-scale, are extracted from the
  \code{HighScaleInput} variable.

@@

In \fs the one-loop electroweak symmetry breaking conditions are
formulated as
%
\begin{align}
  0 = \frac{\partial V^\text{tree}}{\partial v_i} - t_i,
  \label{eq:one-loop-ewsb-eq}
\end{align}
%
where $V^\text{tree}$ is the tree-level Higgs potential, $v_i$ is the
VEVs corresponding to the Higgs fields $H_i$ and $t_i$ is the one-loop
tadpole diagram of $H_i$.

@@

The in more detail the two-scale algorithm used in \fs works as

@@

\subsection{Pole masses}
After the solver routine has finished and convergence has been
achieved, all $\overline{DR}$ parameters are consistent with the
one-loop EWSB conditions, low energy data and all user supplied
boundary conditions are known at any scale between \code{LowScale} and
\code{HighScale}.

@@

For further details, browse the directory \code{examples/tower/}.
One can build and run the example therein by:
\begin{numlstlisting}
$ ./createmodel --name=MSSM
$ ./createmodel --name=MSSMRHN
$ ./configure --with-models=MSSM,MSSMRHN
$ make
$ cd examples/tower
$ make
$ ./run_tower.x --slha-input-file=LesHouches.in.tower
\end{numlstlisting}
In the output,
a part of significant physical interest might be:
\begin{numlstlisting}
Block MSL2 Q=   8.97114431E+02
  1  1     1.25652698E+05   # ml2(1,1)
  1  2    -7.64196328E+01   # ml2(1,2)
  1  3    -7.08429254E+01   # ml2(1,3)
  2  1    -7.64196328E+01   # ml2(2,1)
  2  2     1.25388732E+05   # ml2(2,2)
  2  3    -3.15865242E+02   # ml2(2,3)
  3  1    -7.08429254E+01   # ml2(3,1)
  3  2    -3.15865242E+02   # ml2(3,2)
  3  3     1.24556532E+05   # ml2(3,3)
\end{numlstlisting}
This result demonstrates the well-known effect on
the off-diagonal slepton mass matrix elements
from the non-trivial flavour structure of \code{Yv}
\cite{Borzumati:1986qx}.
This leads in turn to the slepton mixing
matrices, \code{ZE} and \code{ZV},
which contain intergenerational mixings apart from the
generic left-right mixings.

>>>>

  \begin{abstract}
   We introduce \fs, a \mathematica package which generates a fast,
   precise C++ spectrum generator for any SUSY model specified by the
   user.  The generated code is designed with both speed and
   modularity in mind, making it easy to adapt and extend with new
   features. The user specifies the model supplying the
   superpotential, gauge structure and particle content in a \sarah
   model file and provides the specific boundary conditions in a
   separate \fs model file. \fs makes use of the existing \sarah
   package to obtain self energies, tadpole corrections,
   renormalisation group equations (RGEs) and tree level mass and
   electroweak symmetry breaking conditions for the specified model. These are
   translated into C++ code and combined  with numerical
   routines for solving the RGEs, EWSB conditions and simultaneously
   solving for the spectrum consistent with user specified boundary
   conditions.  The modular structure of the generated code allows for
   individual components to be replaced with an alternative if
   available. \fs has been carefully designed to grow as
   alternative solvers and calculators are added.
  \end{abstract}

@@

\section{Program Summary}
\noindent{\em Program title:} \fs\\ {\em Program obtainable from:}
         {\tt http://flexiblesusy.hepforge.org/}\\ {\em Distribution
           format:}\/ tar.gz\\ {\em Programming language:} {\tt
           C++}\\ {\em Computer:}\/ Personal computer\\ {\em Operating
           system:}\/ Tested on Linux 3.x\\ {\em Word size:}\/ 64
         bits\\ {\em External routines:}\/ SARAH 4.0.4, Boost library,
         Eigen, LAPACK\\ {\em
           Typical running time:}\/ 0.1-0.3 seconds per parameter
         point.\\ {\em Nature of problem:}\/ Determining the mass
         spectrum and mixings for any supersymmetric model. The
         generated code must find simultaneous solutions to
         constraints which are specified at two or more different
         renormalisation scales, which are connected by
         renormalisation group equations forming a large set of
         coupled first-order differential equations. \\ {\em Solution method:}\/
         Nested iterative algorithm and numerical minimisation of the
         Higgs potential.  \\ {\em Restrictions:} The couplings must
         remain perturbative at all scales between the highest and
         lowest boundary condition.  \fs~ assumes that all couplings
         of the model are real (i.e.\ $CP-$conserving). Due to the
         modular nature of the generated code adaption and extension
         to overcome restrictions in scope is quite straightforward.

\newpage
\section{Introduction}
Supersymmetry (SUSY) provides the only non-trivial way to extend the
space-time symmetries of the Poincar\'e
group \cite{Coleman:1967ad,Haag:1974qh}, leading many to suspect that
SUSY may be realised in nature in some form. In particular
supersymmetric extensions of the standard model (SM) where SUSY is broken
at the TeV scale have been proposed to solve the hierarchy
problem \cite{Weinberg:1975gm, Weinberg:1979bn, Gildener:1976ai,
  Susskind:1978ms, 'tHooft:1980xb}, allow gauge coupling
unification \cite{Langacker:1990jh, Ellis:1990wk, Amaldi:1991cn,
  Langacker:1991an, Giunti:1991ta} and predict a dark matter candidate
which can fit the observed relic
density \cite{Goldberg:1983nd,Ellis:1983ew}.  Such models have also
been used for baryogenesis or leptogensis to solve the
matter-anti-matter asymmetry of the universe and have been considered
as the low energy effective models originating from string
theory.

@@

Recent experimental developments have also increased the relevancy of
such a tool. From the recent $7$ TeV and $8$ TeV runs at the Large
Hadron Collider (LHC) there have been two important developments.
First low energy signatures expected from such models, such as the
classic jets plus missing transverse energy signature, have not been
observed, substantially raising the lower limit on sparticle masses
(see e.g.~\cite{Aad:2013wta,Chatrchyan:2014lfa}). No other signature
of beyond the standard model (BSM) physics has been observed, leaving
the fundamental questions which motivated BSM physics
unanswered. Secondly ATLAS and CMS discovered \cite{ATLAS:2012ae,
  Chatrchyan:2012tx} a light Higgs of $125$ GeV, within the mass range
that could be accommodated in the MSSM but requiring stops which are
significantly heavier than both the direct collider limits and
indirect limits that appears in constrained models from the
significantly higher limits on first and second generation squarks.

These motivate both exploring non-minimal SUSY models which ameliorate
the naturalness problems by, for example, raising the tree level Higgs
mass, and models developed with a fresh perspective, based on other
considerations.  In both cases exploration of such models can be aided
if it is possible to quickly create a fast spectrum generator.
Currently there is only option for this, a SPheno-like FORTRAN code
which can be generated from
\sarah \cite{Staub:2010ty,Staub:2009bi,Staub:2010jh,Staub:2012pb,Staub:2013tta}.

@@

\paragraph{Alternative boundary value problem solvers}

@@

\fs can be downloaded as a gzipped tar file from the
\url{http://flexiblesusy.hepforge.org/}.  To download and install
version 1.0.0 run:
%
\begin{lstlisting}[language=bash]
$ wget https://www.hepforge.org/archive/flexiblesusy/FlexibleSUSY-1.0.0.tar.gz
$ tar -xf FlexibleSUSY-1.0.0.tar.gz
$ cd FlexibleSUSY-1.0.0/
\end{lstlisting}%% $
%
A CMSSM spectrum generator can be created with the following three
commands:
%
\begin{lstlisting}[language=bash]
$ ./createmodel --name=MSSM
$ ./configure --with-models=MSSM
$ make
\end{lstlisting}%% $
%
The first command creates the model directory \code{models/MSSM/}
together with a CMSSM model file.  The \code{configure} script checks
the system requirements and creates the \code{Makefile}.  See
\code{./configure --help} for more options.  Executing \code{make}
will start \mathematica to generate the spectrum generator and compile
it.  The resulting executable can be run like this:
%
\begin{lstlisting}[language=bash]
$ models/MSSM/run_MSSM.x \
    --slha-input-file=model_files/MSSM/LesHouches.in.MSSM
\end{lstlisting}%% $
%
When executed, the spectrum generator tries to find a set of \DRbar\
model parameters consistent with all CMSSM boundary conditions for the
parameter point given in the SLHA input file
\code{model_files/MSSM/LesHouches.in.MSSM}.  Afterwards, the pole mass
spectrum and mixing matrices are calculated and written to the command
line in SLHA format \cite{Skands:2003cj,Allanach:2008qq}.  For the
parameter point given in the above example the calculated pole mass
spectrum reads
%
\begin{lstlisting}
Block MASS
   1000021     1.15236966E+03   # Glu
   1000024     3.85774334E+02   # Cha_1
   1000037     6.50460073E+02   # Cha_2
        25     1.14766149E+02   # hh_1
        35     7.06792640E+02   # hh_2
        37     7.11388516E+02   # Hpm_2
        36     7.06523105E+02   # Ah_2
   1000012     3.51856376E+02   # Sv_1
   1000014     3.53042556E+02   # Sv_2
   1000016     3.53046504E+02   # Sv_3
   1000022     2.03889780E+02   # Chi_1
   1000023     3.85760714E+02   # Chi_2
   1000025     6.36544884E+02   # Chi_3
   1000035     6.50133768E+02   # Chi_4
   1000001     9.66656018E+02   # Sd_1
   1000003     1.00983181E+03   # Sd_2
   1000005     1.01651873E+03   # Sd_3
   2000001     1.01653005E+03   # Sd_4
   2000003     1.06089534E+03   # Sd_5
   2000005     1.06090238E+03   # Sd_6
   1000011     2.22570305E+02   # Se_1
   1000013     2.29864536E+02   # Se_2
   1000015     2.29888846E+02   # Se_3
   2000011     3.61946671E+02   # Se_4
   2000013     3.61950866E+02   # Se_5
   2000015     3.63136031E+02   # Se_6
   1000002     8.09787818E+02   # Su_1
   1000004     1.01454197E+03   # Su_2
   1000006     1.01981109E+03   # Su_3
   2000002     1.02015269E+03   # Su_4
   2000004     1.05807759E+03   # Su_5
   2000006     1.05808168E+03   # Su_6
\end{lstlisting}

@@

By default, \fs creates three default boundary conditions for each
SUSY model:
%
\begin{itemize}
\item The \emph{high-scale constraint} is indended to set boundary
  conditions on the model parameters at some very high scale, e.g.\
  the GUT scale $M_X$.  The high-scale is defined by the value given
  in the variable \code{HighScale}.  In the model file in
  \secref{Sec:modfile} it is defined to be the unification scale $M_X$
  where $g_1(M_X) = g_2(M_X)$.  The model parameters which, are set in
  the \code{apply()} function at the high-scale, are extracted from the
  \code{HighScaleInput} variable.

@@

In \fs the one-loop electroweak symmetry breaking conditions are
formulated as
%
\begin{align}
  0 = \frac{\partial V^\text{tree}}{\partial v_i} - t_i,
  \label{eq:one-loop-ewsb-eq}
\end{align}
%
where $V^\text{tree}$ is the tree-level Higgs potential, $v_i$ is the
VEV corresponding to the Higgs fields $H_i$ and $t_i$ is the one-loop
tadpole diagram of $H_i$.

@@

In more detail the two-scale algorithm used in \fs works as

@@

\subsection{Pole masses}
After the solver routine has finished and convergence has been
achieved, all $\overline{DR}$ parameters consistent with the
one-loop EWSB conditions, low energy data and all user supplied
boundary conditions are known at any scale between \code{LowScale} and
\code{HighScale}.

@@

For further details, browse the directory \code{examples/tower/}.
One can build and run the example therein by:
\begin{numlstlisting}
$ ./createmodel --name=MSSM
$ ./createmodel --name=MSSMRHN
$ ./configure --with-models=MSSM,MSSMRHN
$ make
$ cd examples/tower
$ make
$ ./run_tower.x --slha-input-file=LesHouches.in.tower
\end{numlstlisting}
In the output,
a part of significant physical interest might be:
\begin{numlstlisting}
Block MSL2 Q=   8.97114431E+02
  1  1     1.25652698E+05   # ml2(1,1)
  1  2    -7.64196328E+01   # ml2(1,2)
  1  3    -7.08429254E+01   # ml2(1,3)
  2  1    -7.64196328E+01   # ml2(2,1)
  2  2     1.25388732E+05   # ml2(2,2)
  2  3    -3.15865242E+02   # ml2(2,3)
  3  1    -7.08429254E+01   # ml2(3,1)
  3  2    -3.15865242E+02   # ml2(3,2)
  3  3     1.24556532E+05   # ml2(3,3)
\end{numlstlisting}
This result demonstrates the well-known effect on
the off-diagonal slepton mass matrix elements
from a non-trivial flavour structure of \code{Yv}
\cite{Borzumati:1986qx}.
This leads in turn to the slepton mixing
matrices, \code{ZE} and \code{ZV},
which contain intergenerational mixings apart from the
generic left-right mixings.
